\boldmath Disconnected pseudo-$C_\ell$ covariances for projected large-scale structure data

### 2019-11-25 draft/preprint_JCAP_v0.2.1-for-arXiv/paper.tex
accepted version of paper in JCAP for arXiv without coloured changes

\abstract{The disconnected part of the power spectrum covariance matrix (also
  known as the ``Gaussian'' covariance) is the dominant contribution on large
  scales for galaxy clustering and weak lensing datasets. The presence of a
  complicated sky mask causes non-trivial correlations between different
  Fourier/harmonic modes, which must be accurately characterized in order to
  obtain reliable cosmological constraints. This is particularly relevant for
  galaxy survey data. Unfortunately, an exact calculation of these
  correlations involves $O(\ell_{\rm max}^6)$ operations that become
  computationally impractical very quickly. We present an implementation of
  approximate methods to estimate the Gaussian covariance matrix of power
  spectra involving spin-0 and spin-2 flat- and curved-sky fields, expanding
  on existing algorithms \cgg{developed in the context of CMB analyses}.
  These methods achieve an $O(\ell_{\rm max}^3)$
  scaling, which makes the computation of the covariance matrix as fast as the
  computation of the power spectrum itself. We quantify the accuracy of these
  methods on large-scale structure and weak lensing data, making use of a
  large number of Gaussian but otherwise realistic simulations. We show that,
  using the approximate covariance matrix, we are able to recover the true
  posterior distribution of cosmological parameters to high accuracy. We also
  quantify the shortcomings of these methods, which become unreliable on the
  very largest scales, as well as for covariance matrix elements involving
  cosmic shear $B$ modes. The algorithms presented here are implemented in the
  public code {\tt NaMaster} \url{https://github.com/LSSTDESC/NaMaster}.}

@@

      Given a field ${\bf a}({\bf x})$, with 1 (spin-0) or 2 components (spin-2), defined on the coordinates ${\bf x}$, we define its generalized Fourier coefficients as
      \begin{equation}
        {\bf a}_{\bf k}=\summ{x}\,\Ylm{\dag}{k}{x}\,{\bf a}({\bf x}),
      \end{equation}
      where the operator $\summ{x}$ denotes an integral or sum over all values of the coordinates ${\bf x}$, and $\Ylm{\dag}{k}{x}$ are a set of orthogonal functions. We will also assume that the $\Ylm{\dag}{k}{x}$ are a \emph{complete} set of basis functions, in which case:
      \begin{align}
        &\summ{x}\,\Ylm{\dag}{k}{x}\Ylm{}{l}{x}=\mathds{1}\Delta^x({\bf k},{\bf l}),\\
        &\summ{k}\,\Ylm{}{k}{x}\Ylm{\dag}{k}{y}=\mathds{1}\Delta^k({\bf x},{\bf y}),
      \end{align}
      where $\summ{k}$ denotes an integral over all possible generalized Fourier coefficients ${\bf k}$, and $\Delta^x$ and $\Delta^k$ are generalized delta functions, defined through their actionl on functions of ${\bf x}$ or ${\bf k}$:
      \begin{align}
        &\summ{k}\,f({\bf k})\Delta^x({\bf k},{\bf l})\equiv f({\bf l}),\\
        &\summ{x}\,f({\bf x})\Delta^k({\bf x},{\bf y})\equiv f({\bf y}).
      \end{align}
      For a spin-$s$ quantity, $\Ylm{}{l}{x}$ can be written in terms of two spin-raising and spin-lowering operators, $\eth$ and $\bar{\eth}$, and a set of scalar orthogonal functions $q({\bf l},{\bf x})$ as:
      \begin{equation}
        \Ylm{}{l}{x}=-\frac{\beta_{\ell,s}}{2}\left(
        \begin{array}{cc}
          \eth^s+\bar{\eth}^s & i(\eth^s-\bar{\eth}^s)\\
          -i(\eth^s-\bar{\eth}^s) & \eth^s+\bar{\eth}^s
        \end{array}\right) q({\bf l},{\bf x}),
      \end{equation}
      where $\beta_{\ell,s}$ is a normalization factor defined in Table \ref{tab:notation}.

@@

      The pseudo-$C_\ell$ estimator then proceeds in two steps:
      \begin{enumerate}
        \item We first bin different ${\bf l}$ modes into sets of them called bandpowers (typically bands of similar $\ell$ or annuli of flat-sky Fourier modes spanning a range of radii). Let us denote a given bandpower by its index $q$. \cgg{We must note that it is more appropriate to use bandpower-averaged spectra when the underlying spectrum does not vary much within each $\ell$ bin. When this is not the case, it is often useful to apply $\ell$-weights (such as $D_\ell\equiv\ell(\ell+1)C_\ell/(2\pi)$). The large-scale structure spectra discussed here are sufficiently flat that the binning used is appropriate.} The binned pseudo-power spectrum is:
        \begin{equation}\label{eq:bandpowers}
          \tilde{\sf C}^{ab}_q=\sum_{{\bf l}\in q} B_q^{\bf l}\,\tilde{\bf a}_{\bf l}\tilde{\bf b}^\dag_{\bf l},
        \end{equation}
        where the bandpower weights are normalized such that $\sum_{{\bf l}\in
          q}B_q^{\bf l}=(K\Delta^x({\bf 0}))^{-1}$.
        \item Then, the correlation between bandpowers induced by the mode-coupling coefficients is partially reversed by multiplying $\tilde{\sf C}^{ab}_q$ by the so-called binned ``mode-coupling matrix'' $\mathcal{M}$, giving the final estimator
        \begin{equation}
          {\rm vec}\left(\hat{\sf C}^{ab}_q\right) = \sum_{q'} \left(\mathcal{M}^{-1}\right)_{qq'}\,{\rm vec}\left(\tilde{\sf C}^{ab}_{q'}\right).
        \end{equation}

@@

      Now, in what follows, we will be concerned with the auto- and
      cross-correlations of spin-0 and spin-2 fields. Thus, to simplify the
      notation, we will enumerate the different types of coupling coefficients
      that exist for a spin-0 field with a single component that we will call
      $\delta$, in analogy to the projected galaxy overensity, and for a
      spin-2 field, $\boldsymbol{\gamma}$, with $E$ and $B$ components, $\gamma_E$ and $\gamma_B$, in analogy to the cosmic shear field. With this setup, all the possible non-zero $W^{\alpha\beta,\alpha'\beta'}_{{\bf l}{\bf l}'}$ can be expressed in terms of the $I^{\pm s_a,\pm s_b}_{{\bf l}{\bf l}'}$ as follows:
      \begin{align}\label{eq:coeff}
        &W^{\delta\delta,\delta\delta}_{{\bf l},{\bf l}'}=I^{0,0}_{{\bf l}{\bf l}'}; \\
        &W^{\delta\gamma_E,\delta\gamma_E}_{{\bf l},{\bf l}'}=W^{\delta\gamma_B,\delta\gamma_B}_{{\bf l},{\bf l}'}=I^{0,+2}_{{\bf l},{\bf l}'};\hspace{12pt}
        W^{\delta\gamma_E,\delta\gamma_B}_{{\bf l},{\bf l}'}=-W^{\delta\gamma_B,\delta\gamma_E}_{{\bf l},{\bf l}'}=-i\,I^{0,-2}_{{\bf l},{\bf l}'};\\
        &W^{\gamma_E\gamma_E,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_E\gamma_B,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_E,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_B,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=I^{+2,+2}_{{\bf l}{\bf l}'}\\
        &W^{\gamma_E\gamma_E,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_B,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_E,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_E\gamma_B,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=I^{-2,-2}_{{\bf l}{\bf l}'}\\
        &W^{\gamma_E\gamma_E,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_E\gamma_B,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_E,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_B,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=-i\,I^{+2,-2}_{{\bf l}{\bf l}'}\\
        &W^{\gamma_E\gamma_E,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_E\gamma_B,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_E,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_B,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=i\,I^{-2,+2}_{{\bf l}{\bf l}'}.
      \end{align}
      Thus, in principle, we only need to compute 7 different types of terms
      ($I^{0,0}$, $I^{0,\pm2}$, $I^{\pm2,\pm2}$ and $I^{\pm2,\mp2})$). In
      order to simplify these expressions further, we follow \cite{Efstathiou:2006eb,2017A&A...602A..41C} and neglect all gradients of the masks\footnote{\cgg{It is worth noting that it is possible in principle to avoid this approximation, as demonstrated in \cite{2005MNRAS.360..509C}}.}, which allows us to relate the different $I^{\pm s_a,\pm s_b}$ through the following set of identities:
      \begin{align}\nonumber
        \summ{x}w({\bf x})\left(\bar{\eth}^2q({\bf l},{\bf x})\right)^*\,\bar{\eth}^2q({\bf k},{\bf x})
        &=\summ{x}w({\bf x})\eth^2 q^*({\bf l},{\bf x})\,\bar{\eth}^2q({\bf k},{\bf x})\\\nonumber
        &=\summ{x}q^*({\bf l},{\bf x})\,\eth^2\left(\bar{\eth}^2q({\bf k},{\bf x})w({\bf x})\right)\\\nonumber
        &\simeq\summ{x}q^*({\bf l},{\bf x})\,\left(\eth^2\bar{\eth}^2q({\bf k},{\bf x})\right)w({\bf x})\\
        &=\frac{1}{\beta^2_{k,2}}\summ{x}q^*({\bf l},{\bf x})\,q({\bf k},{\bf x})w({\bf x})\\\nonumber
        &=\summ{x}q^*({\bf l},{\bf x})\,\left(\bar{\eth}^2\eth^2q({\bf k},{\bf x})\right)w({\bf x})\\\nonumber
        &=\summ{x}\bar{\eth}^2\left(w({\bf x})\,q^*({\bf l},{\bf x})\right)\,\eth^2q({\bf k},{\bf x})\\
        &\simeq\summ{x}w({\bf x})\,\left(\eth^2\,q({\bf l},{\bf x})\right)^*\,\eth^2q({\bf k},{\bf x}),
      \end{align}
      where we have made repeated use of integration by parts.

@@

      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/contaminants_diag_cov.pdf}
        \caption{Diagonal of the covariance matrix estimated from simulations without contaminants (solid blue), with 200 contaminants deprojected (dashed orange) and the NKA estimator (solid green). The differences between the simulated cases are significantly smaller than those associated with the NKA estimator. The NKA method is therefore equally valid to approximate the covariance even in the presence of contaminant deprojection.}
        \label{fig:diag_conts}
      \end{figure}
      Figure \ref{fig:diag_conts} serves also to illustrate another important
      point. In all our tests we find that the largest differences between the
      sample and NKA covariances occur on large scales. For $\delta$-$\delta$
      correlations, the effect is limited to the first few multipoles
      ($\ell\lesssim10$), while for $\gamma_E$-$\gamma_E$ we are only able to
      recover the sample covariance errors within $5\%$ for $\ell\gtrsim40$.
      This is the main source of the small mismatch observed in
      Figure~\ref{fig:chi2_1bin}. Note, however, that most of the cosmological
      information is obtained from the higher multipoles, due to their higher
      statistical weight, and, as we will show in the next section, the effect
      on the final parameter constraints is negligible. If accurate
      \cgg{covariance matrix elements} are needed on these large scales for cosmic shear, they
      can be estimated alternatively making use of fast, low resolution
      simulations (e.g. {\tt HEALPix} $N_{\rm side}=64$ maps)\cgg{, or computed
      exactly as in Ref.~\cite{Efstathiou:2006eb,2017A&A...602A..41C}}.

    \subsection{Impact on parameter estimation}\label{ssec:results.cosmo}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/contours_2bin.pdf}
        \caption{1$\sigma$ and 2$\sigma$ contours for $\Omega_M$ and $\sigma_8$ found from two different estimates of the power spectrum likelihood for a dataset containing all possible cross-power spectra between $\delta$ and $\gamma_E$ for two redshift bins. The filled contours correspond to the parameter likelihood evaluated using the NKA estimator for the power spectrum covariance matrix, while the dashed black contours correspond to the same calculation done with the sample covariance matrix corrected by the factor in Eq. \ref{eq:hartlap}. In both cases we assume a Gaussian likelihood. We find a remarkably good agreement between both likelihoods, highlighting the negligible impact of the approximations involved in the NKA method on the final parameter estimates.} \label{fig:contours}
      \end{figure}
      Ultimately, the most important test to judge the accuracy of the
      analytical covariance matrix estimators implemented here is to study
      their impact on the posterior distribution of cosmological parameters
      derived from power spectrum measurements. Assuming flat priors \cgg{and
      a fiducial Gaussian likelihood approximation~\cite{2008PhRvD..77j3013H}, in which the covariance matrix is computed only once for the fiducial model}, the posterior distribution for parameters $\vec{\theta}$ is simply given by:
      \begin{equation}\label{eq:parpos}
        -2\log p(\vec{\theta}|\hat{\bf C})=\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)^T\cdot{\rm Cov}^{-1}\cdot\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)+{\rm const.},
      \end{equation}
      where $\hat{\bf C}$ is a vector of power spectrum measurements, ${\rm Cov}$ is their covariance matrix, and ${\bf C}(\vec{\theta})$ is their theoretical prediction for parameters $\vec{\theta}$. \cgg{It is worth noting that this Gaussian likelihood is not accurate on large scales, where the small number of modes invalidates the application of the central limit theorem. Even in this regime, the likelihood can be approximated through the method described in \cite{2008PhRvD..77j3013H}, which still requires an accurate estimate of the power spectrum covariance.}

      We explore $\log p(\vec{\theta}|\hat{\bf C})$ for \cgg{the} two parameters $\vec{\theta}\equiv(\Omega_M,\sigma_8)$, for a data vector composed of all possible auto- and cross-correlations between $\delta$ and $\gamma_E$ in the case of two redshift bins described in Section \ref{ssec:results.sims}. In this simple two-dimensional scenario, we simply sample the distribution in a regular grid of 100 by 100 points for each parameter. We construct a data vector $\hat{\bf C}$ from the theoretical prediction for the experimental setup described in Section \ref{ssec:results.sims}, and produce theoretical predictions for it at each grid point using {\tt CCL}. Note that, when evaluating the posterior for the sample covariance matrix, one needs to correct for the finite number of simulations used to construct the covariance. In most situations this can be done simply by rescaling the inverse covariance matrix by a factor given by \cite{2007A&A...464..399H}
      \begin{equation}\label{eq:hartlap}
        {\rm Cov}^{-1}\,\longrightarrow \frac{N_s-2-N_{\rm data}}{N_s-1}\,{\rm Cov}^{-1},
      \end{equation}
      where $N_s=20,000$ is the number of samples used to estimate the
      covariance, and $N_{\rm data}=3510$ is the number of data points.

@@

  \section{Discussion}\label{sec:discussion}
    Estimating accurate covariance matrices for projected two-point
    correlators is an ubiquitous problem in modern cosmology
    \cite{2013PhRvD..88f3537D,2013MNRAS.432.1928T,2017arXiv170609359K}, but it
    is particularly relevant for large-scale structure datasets. This problem
    is further complicated in this case, in comparison with e.g. CMB
    experiments, by two factors: 
    the fact that the fields involved are
    non-Gaussian at some level and the \cgg{arguably} higher complexity of the sky masks used in optical datasets.
    The main impact of survey geometry is the statistical
    coupling it induces between different Fourier/harmonic modes, which must
    be accurately characterized in order to obtain reliable estimates of the
    posterior distribution for cosmological parameters. In this paper we have
    focused on the impact of survey geometry on the dominant Gaussian (i.e.
    disconnected) part of the covariance matrix.

    We have described and generalized existing analytical approaches to
    estimate the covariance matrix for pseudo-$C_\ell$ power spectrum
    estimators
    \cite{2004MNRAS.349..603E,2005MNRAS.360.1262B,2017A&A...602A..41C}, and
    implemented them in the public code {\tt NaMaster}
    \cite{2019MNRAS.484.4127A}, making it straightforward to fully account for
    the effects of survey geometry on the data uncertainties. With these
    approximation\cgg{s}, computationally speaking, the problem of estimating a
    covariance matrix is as complex as that of computing the power spectrum
    itself, and scales with the number of pixels in the map as $N_{\rm
      pix}^{3/2}$. \cgg{We leave for future work the study of the impact of
      extra complications, as position-dependent noise, on their performance.}

>>>>

\abstract{The disconnected part of the power spectrum covariance matrix (also
  known as the ``Gaussian'' covariance) is the dominant contribution on large
  scales for galaxy clustering and weak lensing datasets. The presence of a
  complicated sky mask causes non-trivial correlations between different
  Fourier/harmonic modes, which must be accurately characterized in order to
  obtain reliable cosmological constraints. This is particularly relevant for
  galaxy survey data. Unfortunately, an exact calculation of these
  correlations involves $O(\ell_{\rm max}^6)$ operations that become
  computationally impractical very quickly. We present an implementation of
  approximate methods to estimate the Gaussian covariance matrix of power
  spectra involving spin-0 and spin-2 flat- and curved-sky fields, expanding
  on existing algorithms developed in the context of CMB analyses.
  These methods achieve an $O(\ell_{\rm max}^3)$
  scaling, which makes the computation of the covariance matrix as fast as the
  computation of the power spectrum itself. We quantify the accuracy of these
  methods on large-scale structure and weak lensing data, making use of a
  large number of Gaussian but otherwise realistic simulations. We show that,
  using the approximate covariance matrix, we are able to recover the true
  posterior distribution of cosmological parameters to high accuracy. We also
  quantify the shortcomings of these methods, which become unreliable on the
  very largest scales, as well as for covariance matrix elements involving
  cosmic shear $B$ modes. The algorithms presented here are implemented in the
  public code {\tt NaMaster} \url{https://github.com/LSSTDESC/NaMaster}.}

@@

      Given a field ${\bf a}({\bf x})$, with 1 (spin-0) or 2 components (spin-2), defined on the coordinates ${\bf x}$, we define its generalized Fourier coefficients as
      \begin{equation}
        {\bf a}_{\bf k}=\summ{x}\,\Ylm{\dag}{k}{x}\,{\bf a}({\bf x}),
      \end{equation}
      where the operator $\summ{x}$ denotes an integral or sum over all values of the coordinates ${\bf x}$, and $\Ylm{\dag}{k}{x}$ are a set of orthogonal functions. We will also assume that the $\Ylm{\dag}{k}{x}$ are a \emph{complete} set of basis functions, in which case:
      \begin{align}
        &\summ{x}\,\Ylm{\dag}{k}{x}\Ylm{}{l}{x}=\mathds{1}\Delta^x({\bf k},{\bf l}),\\
        &\summ{k}\,\Ylm{}{k}{x}\Ylm{\dag}{k}{y}=\mathds{1}\Delta^k({\bf x},{\bf y}),
      \end{align}
      where $\summ{k}$ denotes an integral over all possible generalized Fourier coefficients ${\bf k}$, and $\Delta^x$ and $\Delta^k$ are generalized delta functions, defined through their action on functions of ${\bf x}$ or ${\bf k}$:
      \begin{align}
        &\summ{k}\,f({\bf k})\Delta^x({\bf k},{\bf l})\equiv f({\bf l}),\\
        &\summ{x}\,f({\bf x})\Delta^k({\bf x},{\bf y})\equiv f({\bf y}).
      \end{align}
      For a spin-$s$ quantity, $\Ylm{}{l}{x}$ can be written in terms of two spin-raising and spin-lowering operators, $\eth$ and $\bar{\eth}$, and a set of scalar orthogonal functions $q({\bf l},{\bf x})$ as:
      \begin{equation}
        \Ylm{}{l}{x}=-\frac{\beta_{\ell,s}}{2}\left(
        \begin{array}{cc}
          \eth^s+\bar{\eth}^s & i(\eth^s-\bar{\eth}^s)\\
          -i(\eth^s-\bar{\eth}^s) & \eth^s+\bar{\eth}^s
        \end{array}\right) q({\bf l},{\bf x}),
      \end{equation}
      where $\beta_{\ell,s}$ is a normalization factor defined in Table \ref{tab:notation}.

@@

      The pseudo-$C_\ell$ estimator then proceeds in two steps:
      \begin{enumerate}
        \item We first bin different ${\bf l}$ modes into sets of them called bandpowers (typically bands of similar $\ell$ or annuli of flat-sky Fourier modes spanning a range of radii). Let us denote a given bandpower by its index $q$. We must note that it is more appropriate to use bandpower-averaged spectra when the underlying spectrum does not vary much within each $\ell$ bin. When this is not the case, it is often useful to apply $\ell$-weights (such as $D_\ell\equiv\ell(\ell+1)C_\ell/(2\pi)$). The large-scale structure spectra discussed here are sufficiently flat that the binning used is appropriate. The binned pseudo-power spectrum is:
        \begin{equation}\label{eq:bandpowers}
          \tilde{\sf C}^{ab}_q=\sum_{{\bf l}\in q} B_q^{\bf l}\,\tilde{\bf a}_{\bf l}\tilde{\bf b}^\dag_{\bf l},
        \end{equation}
        where the bandpower weights are normalized such that $\sum_{{\bf l}\in
          q}B_q^{\bf l}=(K\Delta^x({\bf 0}))^{-1}$.
        \item Then, the correlation between bandpowers induced by the mode-coupling coefficients is partially reversed by multiplying $\tilde{\sf C}^{ab}_q$ by the so-called binned ``mode-coupling matrix'' $\mathcal{M}$, giving the final estimator
        \begin{equation}
          {\rm vec}\left(\hat{\sf C}^{ab}_q\right) = \sum_{q'} \left(\mathcal{M}^{-1}\right)_{qq'}\,{\rm vec}\left(\tilde{\sf C}^{ab}_{q'}\right).
        \end{equation}

@@

      Now, in what follows, we will be concerned with the auto- and
      cross-correlations of spin-0 and spin-2 fields. Thus, to simplify the
      notation, we will enumerate the different types of coupling coefficients
      that exist for a spin-0 field with a single component that we will call
      $\delta$, in analogy to the projected galaxy overensity, and for a
      spin-2 field, $\boldsymbol{\gamma}$, with $E$ and $B$ components, $\gamma_E$ and $\gamma_B$, in analogy to the cosmic shear field. With this setup, all the possible non-zero $W^{\alpha\beta,\alpha'\beta'}_{{\bf l}{\bf l}'}$ can be expressed in terms of the $I^{\pm s_a,\pm s_b}_{{\bf l}{\bf l}'}$ as follows:
      \begin{align}\label{eq:coeff}
        &W^{\delta\delta,\delta\delta}_{{\bf l},{\bf l}'}=I^{0,0}_{{\bf l}{\bf l}'}; \\
        &W^{\delta\gamma_E,\delta\gamma_E}_{{\bf l},{\bf l}'}=W^{\delta\gamma_B,\delta\gamma_B}_{{\bf l},{\bf l}'}=I^{0,+2}_{{\bf l},{\bf l}'};\hspace{12pt}
        W^{\delta\gamma_E,\delta\gamma_B}_{{\bf l},{\bf l}'}=-W^{\delta\gamma_B,\delta\gamma_E}_{{\bf l},{\bf l}'}=-i\,I^{0,-2}_{{\bf l},{\bf l}'};\\
        &W^{\gamma_E\gamma_E,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_E\gamma_B,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_E,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_B,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=I^{+2,+2}_{{\bf l}{\bf l}'}\\
        &W^{\gamma_E\gamma_E,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_B,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_E,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_E\gamma_B,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=I^{-2,-2}_{{\bf l}{\bf l}'}\\
        &W^{\gamma_E\gamma_E,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_E\gamma_B,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_B\gamma_E,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_B,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=-i\,I^{+2,-2}_{{\bf l}{\bf l}'}\\
        &W^{\gamma_E\gamma_E,\gamma_B\gamma_E}_{{\bf l},{\bf l}'}=W^{\gamma_E\gamma_B,\gamma_B\gamma_B}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_E,\gamma_E\gamma_E}_{{\bf l},{\bf l}'}=-W^{\gamma_B\gamma_B,\gamma_E\gamma_B}_{{\bf l},{\bf l}'}=i\,I^{-2,+2}_{{\bf l}{\bf l}'}.
      \end{align}
      Thus, in principle, we only need to compute 7 different types of terms
      ($I^{0,0}$, $I^{0,\pm2}$, $I^{\pm2,\pm2}$ and $I^{\pm2,\mp2})$). In
      order to simplify these expressions further, we follow \cite{Efstathiou:2006eb,2017A&A...602A..41C} and neglect all gradients of the masks\footnote{It is worth noting that it is possible in principle to avoid this approximation, as demonstrated in \cite{2005MNRAS.360..509C}.}, which allows us to relate the different $I^{\pm s_a,\pm s_b}$ through the following set of identities:
      \begin{align}\nonumber
        \summ{x}w({\bf x})\left(\bar{\eth}^2q({\bf l},{\bf x})\right)^*\,\bar{\eth}^2q({\bf k},{\bf x})
        &=\summ{x}w({\bf x})\eth^2 q^*({\bf l},{\bf x})\,\bar{\eth}^2q({\bf k},{\bf x})\\\nonumber
        &=\summ{x}q^*({\bf l},{\bf x})\,\eth^2\left(\bar{\eth}^2q({\bf k},{\bf x})w({\bf x})\right)\\\nonumber
        &\simeq\summ{x}q^*({\bf l},{\bf x})\,\left(\eth^2\bar{\eth}^2q({\bf k},{\bf x})\right)w({\bf x})\\
        &=\frac{1}{\beta^2_{k,2}}\summ{x}q^*({\bf l},{\bf x})\,q({\bf k},{\bf x})w({\bf x})\\\nonumber
        &=\summ{x}q^*({\bf l},{\bf x})\,\left(\bar{\eth}^2\eth^2q({\bf k},{\bf x})\right)w({\bf x})\\\nonumber
        &=\summ{x}\bar{\eth}^2\left(w({\bf x})\,q^*({\bf l},{\bf x})\right)\,\eth^2q({\bf k},{\bf x})\\
        &\simeq\summ{x}w({\bf x})\,\left(\eth^2\,q({\bf l},{\bf x})\right)^*\,\eth^2q({\bf k},{\bf x}),
      \end{align}
      where we have made repeated use of integration by parts.

@@

      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/contaminants_diag_cov.pdf}
        \caption{Diagonal of the covariance matrix estimated from simulations without contaminants (solid blue), with 200 contaminants deprojected (dashed orange) and the NKA estimator (solid green). The differences between the simulated cases are significantly smaller than those associated with the NKA estimator. The NKA method is therefore equally valid to approximate the covariance even in the presence of contaminant deprojection.}
        \label{fig:diag_conts}
      \end{figure}
      Figure \ref{fig:diag_conts} serves also to illustrate another important
      point. In all our tests we find that the largest differences between the
      sample and NKA covariances occur on large scales. For $\delta$-$\delta$
      correlations, the effect is limited to the first few multipoles
      ($\ell\lesssim10$), while for $\gamma_E$-$\gamma_E$ we are only able to
      recover the sample covariance errors within $5\%$ for $\ell\gtrsim40$.
      This is the main source of the small mismatch observed in
      Figure~\ref{fig:chi2_1bin}. Note, however, that most of the cosmological
      information is obtained from the higher multipoles, due to their higher
      statistical weight, and, as we will show in the next section, the effect
      on the final parameter constraints is negligible. If accurate
      covariance matrix elements are needed on these large scales for cosmic shear, they
      can be estimated alternatively making use of fast, low resolution
      simulations (e.g. {\tt HEALPix} $N_{\rm side}=64$ maps), or computed
      exactly as in Ref.~\cite{Efstathiou:2006eb,2017A&A...602A..41C}.

    \subsection{Impact on parameter estimation}\label{ssec:results.cosmo}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/contours_2bin.pdf}
        \caption{1$\sigma$ and 2$\sigma$ contours for $\Omega_M$ and $\sigma_8$ found from two different estimates of the power spectrum likelihood for a dataset containing all possible cross-power spectra between $\delta$ and $\gamma_E$ for two redshift bins. The filled contours correspond to the parameter likelihood evaluated using the NKA estimator for the power spectrum covariance matrix, while the dashed black contours correspond to the same calculation done with the sample covariance matrix corrected by the factor in Eq. \ref{eq:hartlap}. In both cases we assume a Gaussian likelihood. We find a remarkably good agreement between both likelihoods, highlighting the negligible impact of the approximations involved in the NKA method on the final parameter estimates.} \label{fig:contours}
      \end{figure}
      Ultimately, the most important test to judge the accuracy of the
      analytical covariance matrix estimators implemented here is to study
      their impact on the posterior distribution of cosmological parameters
      derived from power spectrum measurements. Assuming flat priors and
      a fiducial Gaussian likelihood approximation~\cite{2008PhRvD..77j3013H}, in which the covariance matrix is computed only once for the fiducial model, the posterior distribution for parameters $\vec{\theta}$ is simply given by:
      \begin{equation}\label{eq:parpos}
        -2\log p(\vec{\theta}|\hat{\bf C})=\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)^T\cdot{\rm Cov}^{-1}\cdot\left(\hat{\bf C}-{\bf C}(\vec{\theta})\right)+{\rm const.},
      \end{equation}
      where $\hat{\bf C}$ is a vector of power spectrum measurements, ${\rm Cov}$ is their covariance matrix, and ${\bf C}(\vec{\theta})$ is their theoretical prediction for parameters $\vec{\theta}$. It is worth noting that this Gaussian likelihood is not accurate on large scales, where the small number of modes invalidates the application of the central limit theorem. Even in this regime, the likelihood can be approximated through the method described in \cite{2008PhRvD..77j3013H}, which still requires an accurate estimate of the power spectrum covariance.

      We explore $\log p(\vec{\theta}|\hat{\bf C})$ for the two parameters $\vec{\theta}\equiv(\Omega_M,\sigma_8)$, for a data vector composed of all possible auto- and cross-correlations between $\delta$ and $\gamma_E$ in the case of two redshift bins described in Section \ref{ssec:results.sims}. In this simple two-dimensional scenario, we simply sample the distribution in a regular grid of 100 by 100 points for each parameter. We construct a data vector $\hat{\bf C}$ from the theoretical prediction for the experimental setup described in Section \ref{ssec:results.sims}, and produce theoretical predictions for it at each grid point using {\tt CCL}. Note that, when evaluating the posterior for the sample covariance matrix, one needs to correct for the finite number of simulations used to construct the covariance. In most situations this can be done simply by rescaling the inverse covariance matrix by a factor given by \cite{2007A&A...464..399H}
      \begin{equation}\label{eq:hartlap}
        {\rm Cov}^{-1}\,\longrightarrow \frac{N_s-2-N_{\rm data}}{N_s-1}\,{\rm Cov}^{-1},
      \end{equation}
      where $N_s=20,000$ is the number of samples used to estimate the
      covariance, and $N_{\rm data}=3510$ is the number of data points.

@@

  \section{Discussion}\label{sec:discussion}
    Estimating accurate covariance matrices for projected two-point
    correlators is an ubiquitous problem in modern cosmology
    \cite{2013PhRvD..88f3537D,2013MNRAS.432.1928T,2017arXiv170609359K}, but it
    is particularly relevant for large-scale structure datasets. This problem
    is further complicated in this case, in comparison with e.g. CMB
    experiments, by two factors: 
    the fact that the fields involved are
    non-Gaussian at some level and the arguably higher complexity of the sky masks used in optical datasets.
    The main impact of survey geometry is the statistical
    coupling it induces between different Fourier/harmonic modes, which must
    be accurately characterized in order to obtain reliable estimates of the
    posterior distribution for cosmological parameters. In this paper we have
    focused on the impact of survey geometry on the dominant Gaussian (i.e.
    disconnected) part of the covariance matrix.

    We have described and generalized existing analytical approaches to
    estimate the covariance matrix for pseudo-$C_\ell$ power spectrum
    estimators
    \cite{2004MNRAS.349..603E,2005MNRAS.360.1262B,2017A&A...602A..41C}, and
    implemented them in the public code {\tt NaMaster}
    \cite{2019MNRAS.484.4127A}, making it straightforward to fully account for
    the effects of survey geometry on the data uncertainties. With these
    approximations, computationally speaking, the problem of estimating a
    covariance matrix is as complex as that of computing the power spectrum
    itself, and scales with the number of pixels in the map as $N_{\rm
      pix}^{3/2}$. We leave for future work the study of the impact of
      extra complications, as position-dependent noise, on their performance.
